A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-20 03-semantics.tex
Pequenas correções e alguns comentários

We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, and~$\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
share identifiers: any identifier is either a variable, an external event,
or an internal event.

@@

The core of our semantics describes how a program reacts to a single external 
input event, i.e., starting from an input event, how the program behaves and 
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way 
that at most one transition is possible at any time, resulting in deterministic 
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
nonnegative integers representing the current stack level,
and~$e,e'\in\E\cup\{\nil\}$ are the events emitted before and after the
transition (both possibly the empty event~$\nil$).

@@

Rule \R{push} matches whenever there is an emitted event in the input
description,
and instantly broadcasts the event to the program, which means
    (a)~awaking active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ expressions altogether (see function~$\bcast$ in
        Figure~\ref{fig.bcast}),
    (b)~creating a nested reaction by increasing the stack level, and, at the same time, and
    (c)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule in the semantics that matches an
emitted event and also immediately consumes it.

@@

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$, $\ceu{\Fin}$,
and~$\ceu{\CanRun}$ expressions in parallel trails.
%
If no trails are blocked in~$\ceu{\CanRun}$ expressions, it means that the program
cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \R{pop}).

@@

Rule \textbf{pop} either has $p=@nop$ or $isblocked(n,p)$ as preconditions.
%
Note that rule \textbf{pop} only applies syntactically to top-level
transitions.
For instance, it can never match $\NST$ rules for subprograms as in rule
\textbf{seq-adv}.
%
Hence, for the first case, rule \textbf{pop} only applies, and is the only one
to apply, to $nop$ as the whole program (i.e., a $nop$ not surrounded by other
expressions, such as in rule \textbf{seq-nop}).
%
For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
$\NST$ transitions apply with $isblocked(n,p)$ and vice versa.
Except for $@canrun$, there are no $\NST$ transitions for the other blocking
expressions ($awaitExt$, $awaitInt$, $every$, and $fin$).
However, considering the precondition $\LL p,n,\epsilon \RR$,
$isblocked(n,@canrun(n))$ is false.
Hence, given the preconditions for rule \textbf{pop}, no $\NST$ transitions can
occur.
Conversely, if a $\NST$ transition is possible, then $isblocked(n,p)$ must be
false.
Again, except for $@canrun$, all other transitions do not involve blocking
expressions, hence, for these transitions, $isblocked(n,p)$ must be false.
For rule \textbf{canrun}, a transition can only occur if the current stack
level matches $@canrun(n)$.
In this case, $isblocked(n,@canrun(n))$ is false.

Finally, we need to show that $\NST$ transitions are mutually exclusive among
themselves.
%
Note that most rules have unique syntactic prefixes, e.g., $(@nop~@and~q)$
(rule \textbf{and-nop1}) is trivially mutually exclusive with $(@nop~@loop~p)$
(rule \textbf{or-nop1}).
%
The only exceptions are rules \textbf{and-adv1} vs. \textbf{and-adv2}, and
\textbf{or-adv1} vs. \textbf{or-adv2}.
In both cases, we need to show that if the left branch can advance, then it
cannot be blocked and vice-versa, i.e., that 
$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(n,p)$ are mutually
exclusive, which is exactly the same reasoning for rule \textbf{pop} above.

>>>>

We assume that $\ceu{\Mem}$, $\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$ and~$\ceu{\EmitInt}$ expressions do not
share identifiers: any identifier is either a variable, an external event,
or an internal event.

@@

The core of our semantics describes how a program reacts to a single external
input event, i.e., starting from an input event, how the program behaves and
becomes idle again to proceed to a subsequent reaction.
%
We use a set of small-step operational rules, which are built in such a way
that at most one transition is possible at any time, resulting in deterministic
reactions.
%
The transition rules map a triple with a program~$p$, a stack level~$n$, and an
emitted event~$e$ to a modified triple as follows:
\[
  \<p,n,e>\trans\<p',n',e'>\,,
\]
where~
%
\rc{Na BNF o $p$ não deveria ser $\P$~(maiúsculo) tbm?}
%
$p,p'\in\P$ are abstract-language programs, $n,n'\in\N$ are
nonnegative integers representing the current stack level,
and~
%
\rc{Checar: Acho que o conjunto $\E$~não foi previamente definido em
  lugar nenhum}
%
$e,e'\in\E\cup\{\nil\}$
are the events emitted before and after the
transition (both being possibly the empty event~$\nil$).

@@

Rule \R{push} matches whenever there is an emitted event in the input
description,
and instantly broadcasts the event to the program, which means
    (a)~awaking active $\ceu{\AwaitExt}$ or $\ceu{\AwaitInt}$ expressions altogether (see function~$\bcast$ in
        Figure~\ref{fig.bcast}),
    (b)~creating a nested reaction by increasing the stack level, and, at the same time,
    (c)~consuming the event ($e$ becomes~$\nil$).
%
Rule \R{push} is the only rule in the semantics that matches an
emitted event and also immediately consumes it.

@@

A reaction eventually blocks in~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$, $\ceu{\Fin}$,
and~$\ceu{\CanRun}$ expressions in parallel trails.
%
If no trails are blocked in~$\ceu{\CanRun}$ expressions, it means that the program
cannot advance in the current reaction.
%
However, $\ceu{\CanRun}$ expressions can still resume in lower stack indexes and will
eventually resume in the current reaction (see rule \R{pop}).
%
\rc{Não entendi essa discussão}

@@

Rule \textbf{pop} either has $p=@nop$ or $isblocked(p,n)$ as preconditions.
%
Note that rule \textbf{pop} only applies syntactically to top-level
transitions.
For instance, it can never match $\NST$ rules for subprograms as in rule
\textbf{seq-adv}.
%
Hence, for the first case, rule \textbf{pop} only applies, and is the only one
to apply, to $nop$ as the whole program (i.e., a $nop$ not surrounded by other
expressions, such as in rule \textbf{seq-nop}).
%
For the second case, we need to show that given $\LL p,n,\epsilon \RR$, no
$\NST$ transitions apply with $isblocked(p,n)$ and vice versa.
Except for $@canrun$, there are no $\NST$ transitions for the other blocking
expressions ($awaitExt$, $awaitInt$, $every$, and $fin$).
However, considering the precondition $\LL p,n,\epsilon \RR$,
$isblocked(@canrun(n),n)$ is false.
Hence, given the preconditions for rule \textbf{pop}, no $\NST$ transitions can
occur.
Conversely, if a $\NST$ transition is possible, then $isblocked(p,n)$ must be
false.
Again, except for $@canrun$, all other transitions do not involve blocking
expressions, hence, for these transitions, $isblocked(p,n)$ must be false.
For rule \R{can-run}, a transition can only occur if the current stack
level matches $@canrun(n)$.
In this case, $isblocked(@canrun(n),n)$ is false.

Finally, we need to show that $\NST$ transitions are mutually exclusive among
themselves.
%
Note that most rules have unique syntactic prefixes, e.g., $(@nop~@and~q)$
(rule \textbf{and-nop1}) is trivially mutually exclusive with $(@nop~@loop~p)$
(rule \textbf{or-nop1}).
%
The only exceptions are rules \textbf{and-adv1} vs. \textbf{and-adv2}, and
\textbf{or-adv1} vs. \textbf{or-adv2}.
In both cases, we need to show that if the left branch can advance, then it
cannot be blocked and vice-versa, i.e., that
$\LL p,n,\epsilon \RR \NST \LL p',n,e \RR$ and $isblocked(p,n)$ are mutually
exclusive, which is exactly the same reasoning for rule \textbf{pop} above.
